---
title: 'Week 1 exercises'
author: "Daniel Alonso"
date: "November 18th, 2020"
output: 'pdf_document'
---

Importing libraries

```{r, echo=TRUE, warning=FALSE}
library(ggplot2)
library(foreach)
library(dplyr)
```


## Exercise 1

Simulating 100 trajectories of length $n = 1000$ for $X$ and $Y$.

```{r, echo=TRUE, warning=FALSE}
Traj_X <- data.frame()
Traj_Y <- data.frame()
for (k in 1:100) {
    X <- data.frame(x=1:1000,val=rep(0,1000),run=rep(k,1000)) 
    Y <- data.frame(x=1:1000,val=rep(0,1000),run=rep(k,1000))
    X_r <- rep(0,1000)
    Y_r <- rep(0,1000)
    for (i in 1:1000) {
        if (i > 1) {
            X_r[i] <- 0.5*X_r[i-1]+rnorm(1)
            Y_r[i] <- 2*Y_r[i-1]+rnorm(1)
        }
    }
    X$val <- X_r
    Y$val <- Y_r
    Traj_X <- rbind(Traj_X, X)
    Traj_Y <- rbind(Traj_Y, Y)
}
```

\newpage

### a - Plotting simulated trajectories

#### Trajectories X

```{r, echo=TRUE, warning=FALSE, eval=FALSE}
ggplot(data = Traj_X, aes(x=x, y=val)) + 
    geom_line(aes(colour=run), show.legend=FALSE)
```

#### Trajectories Y

The values of Y are too large for ggplot to show. 

### b - Use the simulated trajectories to estimate the mean and the covariance functions

```{r, echo=TRUE, warning=FALSE}
mean_X = rep(0,100)
mean_Y = rep(0,100)
for (i in 1:100) {
    x = Traj_X %>% filter(run == i) %>% select(val)
    y = Traj_Y %>% filter(run == i) %>% select(val)
    mean_X[i] = mean(x$val)
    mean_Y[i] = mean(y$val)
}
```

The mean of all trajectories of X is the following:

```{r, echo=TRUE, warning=FALSE}
mean(mean_X)
```

The mean of all trajectories of Y is the following:

```{r, echo=TRUE, warning=FALSE}
mean(mean_Y)
```

The mean of the covariances of all combinations of trajectories is the following:

```{r, echo=TRUE, warning=FALSE, eval=FALSE}
covs = rep(0,100*100)
cnt = 0
for (i in 1:100) {
    for (j in 1:100) {
        x = Traj_X %>% filter(run == i) %>% select(val)
        y = Traj_Y %>% filter(run == j) %>% select(val)
        cnt = cnt + 1
        covs[cnt] = cov(x,y)
    }
}
mean(covs)
```

### c - Is the process stationary? Is it weakly stationary?

X is stationary but Y is not

### d - If the process is weakly stationary, use the function acf to display the autocorrelation function and compare with your own estimate

```{r, echo=TRUE, warning=FALSE, eval=FALSE}
acf(X[1,])
L=30
acfx=numeric(length=L)
for (i in 1:L){
  acfx[i]=mean(diag(gX[1:(n+1-i),(i+1):(n+1)]))
}
plot(1:L, acfx)
```

## Exercise 2

$I_{0} = 5$ and $S_{0} = 1000$

$\alpha = 0.0005$

```{r, echo=TRUE, warning=FALSE}
alpha <- 0.004
cases <- c(5)
s <- 1000
i <- 5
while (i > 0) {
    p <- 1 - (1 - 0.0005)^(i)
    i <- rbinom(1, s, p)
    cases <- c(cases, i)
}
```

```{r, echo=TRUE, warning=FALSE}
plot(cases)
```

```{r, echo=TRUE, warning=FALSE}
P <- c(0.1, 0.4, 0.2, 0.3,
       0.6,   0, 0.2, 0.2,
       0.1, 0.3, 0.3, 0.3,
       0.5, 0.3, 0.1, 0.1)
P <- matrix(P,nrow=4,byrow=T)
alpha=rep(1/4,4) #uniform initial distribution
```





## Exercise 4

Given the transition matrix and initial distribution:

```{r, echo=TRUE, warning=FALSE}
P <- c(0.1, 0.4, 0.2, 0.3,
       0.6,   0, 0.2, 0.2,
       0.1, 0.3, 0.3, 0.3,
       0.5, 0.3, 0.1, 0.1)
P <- matrix(P,nrow=4,byrow=T)
alpha=rep(1/4,4) #uniform initial distribution
```

We define the following function to calculate matrix powers (thanks profe!):

```{r, echo=TRUE, warning=FALSE}
matrixpower <- function(M,k) {
  # ARGUMENTS:
  # M: square matrix 
  # k: exponent
  if(dim(M)[1]!=dim(M)[2]) return(print("Error: matrix M is not square"))
  if (k == 0) return(diag(dim(M)[1]))  # if k=0 returns the identity matrix
  if (k == 1) return(M)
  if (k > 1)  return(M %*% matrixpower(M, k-1)) # if k>1 recursively apply the function
}
```


### a - $P(X_{4} = 3 | X_{3} = 4)$

```{r, echo=TRUE, warning=FALSE}
P[4,3]
```

### b - $P(X_{4} = 3 | X_{2} = 4)$

```{r, echo=TRUE, warning=FALSE}
matrixpower(P,2)[4,3]
```

### c - $P(X_{1} = 1)$

```{r, echo=TRUE, warning=FALSE}
(t(alpha)%*%P)[1]
```

### d - $E[X_{2}]$

```{r, echo=TRUE, warning=FALSE}

```