---
title: 'Week 2 exercises'
author: "Daniel Alonso"
date: "November 19th, 2020"
output: 'pdf_document'
---

Importing libraries

```{r, echo=TRUE, warning=FALSE}
library(ggplot2)
library(matlib)
```

We define the following function to calculate matrix powers (thanks profe!):

```{r, echo=TRUE, warning=FALSE}
matrixpower <- function(M,k) {
  # ARGUMENTS:
  # M: square matrix 
  # k: exponent
  if(dim(M)[1]!=dim(M)[2]) return(print("Error: matrix M is not square"))
  if (k == 0) return(diag(dim(M)[1]))  # if k=0 returns the identity matrix
  if (k == 1) return(M)
  if (k > 1)  return(M %*% matrixpower(M, k-1)) # if k>1 recursively apply the function
}
```


## Exercise 1

```{r, echo=TRUE, warning=FALSE}
P <- c(0.1, 0.4, 0.5,
       0.4, 0.6,   0,
       0.6,   0, 0.4)
P <- matrix(P,nrow=3,byrow=T)
```

First we solve $\pi P = \pi$:

$\pi (P-I) = (1,0, \dots, 0)$

$(\pi_{1}, \pi_{2}, \pi_{3}) \begin{pmatrix} 1 & 0.4 & 0.5 \\ 1 & -0.4 & 0 \\ 1 & 0 & -0.6 \end{pmatrix} = (1,0,0)$

Solving the system we get:

$\pi = (\frac{6}{17}, \frac{15}{17}, \frac{10}{17})$

Which is our stationary distribution.

\newpage

## Exercise 2

```{r, echo=TRUE, warning=FALSE}
stationary_dist <- function(P) {
    dim = sqrt(length(P))
    mat = matrix(P,nrow=dim, byrow=T)
    A = mat - diag(dim)
    b = c(1,rep(0,dim-1))
    A[,1] <- rep(1,dim)
    print("The system is the following:")
    showEqn(A, b)

    print("The solution is the following:")
    pi <- matlib::Solve(A, b, fractions = TRUE)
    return(pi)
}

P <- c(0.1, 0.4, 0.5,
       0.4, 0.6,   0,
       0.6,   0, 0.4)

stationary_dist(P)
```

## Exercise 3

```{r, echo=TRUE, warning=FALSE}
P <- c(  0, 1/3,   0, 2/3,
       2/3,   0,   0, 1/3,
         0,   0,   1,   0,
       1/3, 2/3,   0,   0 )
P <- matrix(P,nrow=4,byrow=T)
```

### a - Find the communication classes and classify the states.

We have two communication classes

### b - Find the set of stationary distributions.

First we solve $\pi P = \pi$:

$\pi (P-I) = (1,0, \dots, 0)$

$\begin{pmatrix} \pi_{1} & \pi_{2} & \pi_{3} & \pi_{4} \end{pmatrix} \begin{pmatrix} 1 & \frac{1}{3} & 0 & \frac{2}{3} \\ 1 & -1 & 0 & \frac{1}{3} \\ 1 & 0 & 0 & 0 \\ 1 & \frac{2}{3} & 0 & -1 \end{pmatrix} = (1,0,0,0)$

Solving the system we get:

$\pi = (0, 0, \pi_{3}, 0)$

Which is our stationary distribution, where $\pi_{3}$ is a free variable.

### c - How is the long-run behavior of the chain? Analyze $\lim_{n\to\infty} P^{n}$ with R

```{r, echo=TRUE, warning=FALSE}
matrixpower(P,50)
```

Given that there's 2 classes in this MC, our long run behavior shows the same values for columns and rows in class 1 (nodes 1,2,4) but not for class 2 (node 3).

### d - Does this chain have a limiting distribution?

No, this chain does not have a limiting distribution because $\lim_{n\to\infty} \alpha P^{n} \neq \lambda$