---
title: 'Stochastic Processes: Assignment 1'
author: "Group 1: Javier Esteban Aragoneses, Mauricio Marcos Fajgenbaun, Danyu Zhang, Daniel Alonso"
date: "November 27th, 2020"
output: 'pdf_document'
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "./figures/"
)
```

Importing libraries

```{r, echo=FALSE, warning=FALSE, results='hide'}
library(markovchain)
library(matlib)
library(nleqslv)
library(rootSolve)
library(stringi)
library(randomcoloR)
library(dplyr)
```

# Problem 1

```{python}
import numpy as np
from copy import deepcopy
import pandas as pd
from nltk.corpus import words
import matplotlib.pyplot as plt
import threading as th

freq =  np.loadtxt('./data/Englishcharacters.txt', usecols=range(27))

def transf_minmax(table,a,b):
    return (b-a)*((table - np.min(table))/(np.max(table) - np.min(table))) + a 

def transf_log(table):
    return np.log(table + 1)

freq = transf_log(freq)

with open('./data/messages.txt') as f:
    message = f.readlines()[9].replace('\n','')

def decode(message, freqs, iters):
    # messages per iteration
    msg_iters = {}
    # identity function
    letters = ["a", "b", "c", "d",
                "e", "f", "g", "h", 
                "i", "j", "k", "l", 
                "m", "n", "o", "p", 
                "q", "r", "s", "t", 
                "u", "v", "w", "x", 
                "y", "z"]
    init_letters = deepcopy(letters)
    cd = {l:l for l,l in zip(init_letters,letters)}
    cd.update({' ':' '})
    def f(c):
        return cd[c]
    
    # mapping letters to cols
    fvals = np.array([x for x in range(len(letters))])
    cd_map = {l:v for l,v in zip(letters,fvals)}
    cd_map.update({' ':26})
    def f_map(c):
        return cd_map[c]

    # score function
    def score(fun):
        p = 0
        for i in range(1,len(msg)):
            p = p + freq[f_map(fun(msg[i-1])),f_map(fun(msg[i]))]
        return p
    
    # msg to list
    msg = list(message)

    # letters to modify the cipher
    letters_n = deepcopy(letters)

    for i in range(iters):

        # randomly choose 2 numbers and replace the 2 chosen 
        # vals in a copy of letters
        ch1 = np.random.randint(1,len(letters))
        ch2 = np.random.randint(1,len(letters))
        plc1 = deepcopy(letters_n[ch1])
        plc2 = deepcopy(letters_n[ch2])
        letters_n[ch1] = plc2
        letters_n[ch2] = plc1

        cd_n = {l:v for l,v in zip(init_letters,letters_n)}
        cd_n.update({' ':' '})
        
        # f asterisk
        def f_n(c):
            return cd_n[c]

        #loop with scores
        scr_f = score(f)
        scr_fn = score(f_n)
        scr = scr_fn/scr_f

        # print(f'score f*: {scr_fn}, score f: {scr_f}, ratio: {scr}')
        cond = np.random.rand() <= scr

        if cond:
            letters = deepcopy(letters_n)
            cd = {l:v for l,v in zip(init_letters,letters)}
            cd.update({' ':' '})
            for k in range(len(msg)):
                msg[k] = f(msg[k])
            msg_iters[i] = (scr,''.join([x for x in msg]))
        else:
            for k in range(len(msg)):
                msg[k] = f(msg[k])
        msg = list(message)
    
    # put in a dataframe
    df = {'iter':[it for it in msg_iters.keys()],
          'score':[msg[0] for msg in msg_iters.values()],
          'msg':[msg[1] for msg in msg_iters.values()]}
    return pd.DataFrame(df)

result = decode(message,freq, 50000)
plt.plot(result.sort_values('score')['score'].reset_index(drop=True))
print(result[result['score'] == max(result['score'])])
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
freq <- read.table('./data/Englishcharacters.txt')
freq <- log(freq + 2)
#scaling freq
freq <- ((freq - min(freq))/(max(freq) - min(freq))) + 1
message <- paste(readLines("./data/messages.txt"))[1]

# decode function
decode <- function(message, freqs, iters) {
    # message separated by "" to iterate
    msg <- strsplit(message, "")[[1]]

    # identity function
    fvals <- 1:27
    letters <- c("a", "b", "c", "d",
                "e", "f", "g", "h", 
                "i", "j", "k", "l", 
                "m", "n", "o", "p", 
                "q", "r", "s", "t", 
                "u", "v", "w", "x", 
                "y", "z", " ")
    names(fvals) <- letters
    f <- function(c) return(fvals[c])

    # score function
    score <- function(fun) {
        p <- 1
        for (i in 2:length(msg)) {
            p <- p * freq[fun(msg[i-1]),fun(msg[i])]
        } 
        return(p)
    }

    for (i in 1:iters) {
        # values for the function
        fvals_n <- 1:27

        # randomly choose 2 numbers and replace the 2 chosen vals in a copy of letters
        choice_1 <- as.integer(runif(1, min=1, max=27))
        choice_2 <- as.integer(runif(1, min=1, max=27))
        letters_n <- letters
        letters_n[choice_1] <- letters[choice_2]
        letters_n[choice_2] <- letters[choice_1]
        names(fvals_n) <- letters_n

        # f asterisk
        f_n <- function(c) return(fvals_n[c])
        scr <- score(f_n)/score(f)
        rnd <- runif(1)
        cond <- rnd <= scr
        if (cond == TRUE){
            letters <- letters_n
            names(fvals) <- letters
            f <- function(c) return(fvals[c])
            for (l in 1:length(msg)) {
                msg[l] <- names(f(msg[l]))
            }
        } else if (cond == FALSE){
            for (l in 1:length(msg)) {
                msg[l] <- names(f(msg[l]))
            }
        }

    }
    return(stri_paste(msg,collapse=''))
}
decode(message,freq,100)
```


# Problem 2

# (a)

Let **$N(t)$** be the number of cars arriving at a parking lot by time **$t$**, according to the proposed scenario, we can model **$N(t)$** as a non-homogenous Poisson process. Such process has almost the same process as any other Poisson process, however, its rate is a function of time.

$N(t), t \in [0, \infty)$ is the non-homogenous Poisson process with rate $\lambda (t)$ where:

- $N(0) = 0$
- $N(t)$ has independent increments

We define 8:00 as $t=0$ with the following integrable function and each unit of $t$ equals to 1 hour:

$\lambda (t) = \begin{cases} 100 & 0 \leq t \leq \frac{1}{2} \\
600t - 200 & \frac{1}{2} < t \leq \frac{3}{4} \\
400t - 50 & \frac{3}{4} < t \leq 1 \\
-500t + 850 & 1 < t \leq 1.5 \\
\end{cases}$

So,

$E[N(t)] = \begin{cases} 
\int_{0}^t 100\,dt = 100t & 0 \leq t \leq \frac{1}{2} \\
\int_{\frac{1}{2}}^t 600t - 200 \,dt + 50 = 300(t^2 - \frac{1}{4}) - 200(t - \frac{1}{2}) + 50 & \frac{1}{2} < t \leq \frac{3}{4} \\
\int_{\frac{3}{4}}^t 400t - 50 \,dt + 93.75 = 25(8t^2 - 2t - 3) + 93.75 & \frac{3}{4} < t \leq 1 \\
\int_{1}^t -500t + 850\,dt + 168.75 = -50(5t^2 - 17t + 12) + 168.75 & 1 < t \leq 1.5
\end{cases}$

Given that there is a limit of 150 vehicles:

$E[N(t)] = \begin{cases} 
100t & 0 \leq t \leq \frac{1}{2} \\
300(t^2 - \frac{1}{4}) - 200(t - \frac{1}{2}) + 50 & \frac{1}{2} < t \leq \frac{3}{4} \\
25(8t^2 - 2t - 3) + 93.75 & \frac{3}{4} < t < 0.94468  \\
150 & t \geq 0.94468
\end{cases}$

# (b)

```{r, echo=FALSE, warning=FALSE, message=FALSE, results='hide', fig.height=5.5, fig.width=10}
# trying 0-200
tt = c()
aa = seq(0,200,1)
for (i in aa) {
    tt = c(tt, ppois(149,lambda=i))
}
test <- data.frame(val=aa, tt=tt)
plot(test$val, test$tt, main='Probando entre 0 y 200')
# trying 139-140
tt = c()
aa = seq(139,140,0.01)
for (i in aa) {
    tt = c(tt, ppois(149,lambda=i))
}
test <- data.frame(val=aa, tt=tt)
plot(test$val, test$tt, main='Probando entre 139 y 140')
```

Luego de hacer las pruebas para $\lambda (t)$ obtenemos lo siguiente:

```{r, echo=TRUE, warning=FALSE, message=FALSE}
lambda = 139.6
t = 0.91232
# 8:44 AM
```

Por lo que $t = 0.91232$ horas (aproximadamente a las 8:54 de la mañana).

# (c)

The following function simulates a non-homogenous poisson process from a homogenous poisson process:

```{r, echo=TRUE, warning=FALSE, message=FALSE}
non_hom_poisson <- function(fun,l,a,b,start=0) {
    # This function generates a non-homogenous poisson 
    # process from a homogenous poisson process
    # PARAMS:
    # fun:   if the non-homogenous poisson process has
    #        multiple functions per time subinterval
    #        this parameters represents such function
    # l:     lambda for the homogenous poisson process
    # a:     lower bound for the time subinterval
    # b:     upper bound for the time subinterval
    # start: this parameter is used to keep track of 
    #        the process count.
    
    # We generate the homogenous poisson process 
    # arrival times
    val <- rpois(1,l*(b-a))
    intervals <- (b-a) * sort(runif(val)) + a

    # Non-homogenous poisson process
    evs <- length(intervals) # lenght of arrival times 
    nh_val <- 0 + start # start of the event count
    nh_intervals <- c() # arrival times for the NHPP
    for (i in 1:evs) {
        if (runif(1) < fun(intervals[i])/l) { 
            # only including intervals from the HPP which
            # match with fun(intervals[i])/l probability
            nh_intervals <- c(nh_intervals, intervals[i]) 
            nh_val <- nh_val+1 # adding one to the event count
        }
    }
    nh_events <- seq(1+start,nh_val,1) # events since the previous group
    return(list(arrival_times=nh_intervals, events=nh_events))
}
```

\newpage

```{r, echo=TRUE, warning=FALSE, message=FALSE}
simulation <- function(iters, functions, lambdas, ints) {
    # This function simulates from the NHPP
    # iters:     number of iterations to plot and add to the list of
    #            dataframes
    # functions: list of functions corresponding to the lambda function
    # lambdas:   list of lambdas for each subinterval
    # ints:      lists of vectors of 2 elements each containing the intervals
    #            that correspond to each element of lambdas and functions lists
    p <- list()
    for (i in 1:iters) {
        maximum <- 0 # start for the next NHPP simulation to continue count
        arr_times <- c() # arrival times
        events <- c() # event counts
        for (k in 1:4) {
            int <- non_hom_poisson(lambda_funs[[k]],lambdas[[k]],
                                   ints[[k]][1],ints[[k]][2],
                                   start=maximum)
            maximum <- max(int$events) # remembering last event count
            arr_times <- c(arr_times, int$arrival_times) 
            events <- c(events, int$events)
        }
        p[[i]] <- data.frame(arrival_times=arr_times, events=events)
        # plots
        if (i == 1) {plot(arr_times, events, cex=0.5, pch='.', 
                          col=randomColor(), xlim=c(0,1.5), 
                          ylim=c(0,150))}
        lines(arr_times, events, col=randomColor())
    }
    return(p)
}
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# subintervals
int1 <- c(  0, 1/2)
int2 <- c(1/2, 3/4)
int3 <- c(3/4,   1)
int4 <- c(  1, 1.5)
ints <- list(int1, int2, int3, int4)

# lambdas
lambda_1 <- 100
lambda_2 <- 250
lambda_3 <- 350
lambda_4 <- 350
lambdas <- list(lambda_1, lambda_2, lambda_3, lambda_4)

# lambda functions per t subinterval
l1 <- function(t){100}
l2 <- function(t){600*t-200}
l3 <- function(t){400*t-50}
l4 <- function(t){-500*t+850}
lambda_funs <- list(l1, l2, l3, l4)
```

\newpage

```{r, echo=TRUE, warning=FALSE, message=FALSE, fig.height=16, fig.width=14}
data <- simulation(1000, lambda_funs, lambdas, ints)
```

```{r, echo=TRUE, warning=FALSE, message=FALSE}
ratio <- 0
for (i in 1:length(data)) {
    df <- data.frame(data[[i]])
    cnt <- df %>% filter(arrival_times < 0.91232 & events >= 150) %>% dplyr::count()
    if (cnt[1] >= 1) {
        ratio <- ratio + 1
    }
}
ratio/1000
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
tt = c()
aa = seq(0,140,1)
bb = seq(0, 30,1)
for (i in bb) {
    for (j in aa) {
        tt = c(tt, ppois(149,lambda=i))
    }
}
test <- data.frame(val=aa, tt=tt)
write.csv(test, 'test.csv')
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
test <- 0
for (i in 0:143) {
    test <- test + (exp(-140) * 140^i)/(factorial(i))
}
```

# Problem 3

# (a)

Our infinitesimal generator is the following:

$Q = \begin{pmatrix}
- \lambda & \lambda & 0 & 0 & \dots \\
\mu & - (\lambda + \mu) & \lambda & 0 & \dots \\
0 & 2 \mu & - (2 \mu + \lambda) & \lambda & \dots \\
\vdots & \vdots & \vdots & \vdots & \ddots
\end{pmatrix}$

# (b)

We solve the following system:

$\begin{cases}
\sum_{i = 1}^{\infty} \pi_i = 1 \\
\lambda \pi_0 = \mu \pi_1 \\
\lambda \pi_1 = 2 \mu \pi_2 \\
\vdots \\
\lambda \pi_{n-1} = 2 \mu \pi_n \\
\vdots \\
\end{cases}$

First we have:

$\pi_1 = \frac{\lambda \pi_0}{\mu} \\
\pi_2 = \frac{\lambda^2 \pi_0}{2 \mu^2} \\
\pi_3 = \frac{\lambda^3 \pi_0}{2^2 \mu^3} \\
\vdots \\
\pi_n = \frac{\lambda^n \pi_0}{2^{n-1} \mu^n} \\
\vdots$

Then:

$\sum_{i=0}^{\infty} \pi_i = \pi_0 + \frac{\lambda \pi_0}{\mu} + \frac{\lambda^2 \pi_0}{2 \mu^2} + \dots + \frac{\lambda^n \pi_0}{2^{n-1} \mu^n} + \dots = 1$

And so factoring $\pi_0$ we get:

$\pi_0 (1 + \frac{\lambda}{\mu} + \frac{\lambda^2}{2 \mu^2} + \dots) = \pi_0 (1 + \sum_{i=1}^{\infty} \frac{1}{2^{i-1}} (\frac{\lambda}{\mu})^{i})$

Then multiplying $\frac{2}{2}$ to the summation:

$\pi_0 (1 + \frac{2}{2} \sum_{i=1}^{\infty} \frac{1}{2^{i-1}} (\frac{\lambda}{\mu})^{i})$

= $\pi_0 (2 \sum_{i=0}^{\infty} (\frac{\lambda}{2 \mu})^{i} - 1)$

$\pi_0 (2 (\frac{1}{1 - \frac{\lambda}{2 \mu}}) - 1) = 1$

$\pi_0 = \frac{1}{2 (\frac{1}{1 - \frac{\lambda}{2 \mu}}) - 1}$

$\vdots$

$\pi_n = \frac{\lambda^n}{2^{n-1}{\mu_n} \pi_0} \frac{1}{2 (\frac{1}{1 - \frac{\lambda}{2 \mu}}) - 1}$

finally:

$\pi_n = \frac{1}{2^{n-1}} (\frac{\lambda}{\mu})^{n} \pi_0$

The infinite sum converges when $|\frac{\lambda}{2 \mu}| < 1$ in which case the stationary distribution P exists.

Then:

$L = \sum_{n=0}^{\infty} \pi_n * n$

Using the following sum:

$\sum_{i=0}^{n-1} i a^i = \frac{a - na^n + (n-1)a^{n+1}}{(1-a)^2}$

As n approaches infinity:

$\sum_{i=0}^{\infty} i a^i = \frac{a}{(1-a)^2}$

We get the following:

$L = \pi_0 [1 + \sum_{n=0}^{infty} (\frac{\lambda}{\mu})^n * \frac{n}{2^{n-1}}]$

$L = \pi_0 [1 + 2 \sum_{n=0}^{infty} (\frac{\lambda}{2 \mu})^n * n]$

$L = \pi_0 [1 + 2 \frac{\frac{\lambda}{2 \mu}}{(1 - \frac{\lambda}{2 \mu})^2}]$

# (c)

Let's consider the probabilities conditioned on the number of customers in the system that are present once our specific subject *l* gets into the system.

If there are no other customers when *l* gets into the system, there is no chance of overtaking.

$P(N^{OV} = 0 | N^{PR} = 0) = 1$

With $N^{OV}$ being the number of customers that *l* overtakes and $N^{PR}$ the number of customers present in the system (queing) when *l* gets in the system.

If $N^{PR} = 1$, then *l* can overtake only 1 customer, if the time it takes to be served is shorter than the time it takes the other customers to be served. Because of the memoryless property we can assert the following:

$P(N^{OV} = 0 | N^{PR} = 1) = \frac{\mu}{\mu + \mu} = \frac{1}{2}$

Actually, in general:

$P(N^{OV} = k | N^{PR} = n) = \frac{1}{n+1}$, $n \leq c - 1$, $x = 0,1$

As in this case c=2, our *l* subject can' t overtake more than one customer.

Now, if $n \geq c$, that is, *l* has to get in queue and wait to be served. When *l* gets served, there is also one more customer getting served. Because, again, of the memoryless property.

$P(N^{OV} = k | N^{PR} = n) = \frac{1}{c}$, $n = c$, $k = 0,1$

In our case, it does not matter how many customers are in the system, the probability of overtaking, conditioned to the number of customers already in the system, is $\frac{1}{2}$.

Now, using Bayes' theorem and the total probability rule, we can find the probability of *l* overtaking another customer.

$P(A | B) = \frac{P(A \cap B)}{P(B)}$

$\frac{1}{2} \sum_{i=1}^{\infty} \pi_i = \frac{1}{2} \sum_{i=1}^{\infty} (\frac{1}{2^{i-1}}) (\frac{\lambda}{\mu})^{i} \pi_0$

= $\sum_{i=1}^{\infty} (\frac{\lambda}{2 \mu})^i \pi_0 = (\sum_{i=0}^{\infty} ((\frac{\lambda}{2 \mu})^i) - 1) \pi_0$

= $(\frac{1}{1-\frac{\lambda}{2 \mu}} - 1) \pi_0 = \frac{1}{2 (\frac{1}{1 - \frac{\lambda}{2 \mu} - 1})} (\frac{1}{1 - \frac{\lambda}{\mu}} - 1)$


= $\frac{1}{1 - \frac{\lambda}{2 \mu}} - 1 = \frac{1 - (1 - \frac{\lambda}{2 \mu})}{1 - \frac{\lambda}{2 \mu}} = \frac{\frac{\lambda}{2 \mu}}{1 - \frac{\lambda}{2 \mu}}$


= $2 (\frac{1}{1 - \frac{\lambda}{2 \mu}}) - 1 = \frac{2}{1 - \frac{\lambda}{2 \mu}} - 1 = \frac{2 - (1 - \frac{\lambda}{2 \mu})}{1 - \frac{\lambda}{2 \mu}}$

= $\frac{1 + \frac{\lambda}{2 \mu}}{1 - \frac{\lambda}{2 \mu}}$

Then:

$\frac{\frac{\lambda}{2 \mu}}{1 \frac{\lambda}{2 \mu}} = \frac{\frac{\lambda}{2 \mu}}{\frac{2 \mu + \lambda}{2 \mu}} = \frac{\lambda}{2 \mu + \lambda}$

So then we get:

$P(N^{OV} = k) = \frac{\lambda}{2 \mu + \lambda}$, $k = c-1 = 1$

# (d)

We define the following function to simulate the queueing system:

```{r, echo=TRUE, warning=FALSE, message=FALSE, eval=FALSE}
q <- function(n, l, m) {
    s = cumsum(rexp(n, l))
    t = rexp(n, m)

    exit_time = numeric(n)
    exit_time[1:2] = t[1:2]
    for (i in 3:n) {
        s_t <- sort(exit_time[1:(i-1)], decreasing = TRUE)

        if (s[i])
    }
}
```