---
title: 'Stochastic Processes: Assignment 1'
author: "Group 1: Javier Esteban Aragoneses, Mauricio Marcos Fajgenbaun, Danyu Zhang, Daniel Alonso"
date: "November 27th, 2020"
output: 'pdf_document'
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "./figures/"
)
```

Importing libraries

```{r, echo=FALSE, warning=FALSE, results='hide'}
library(markovchain)
library(matlib)
library(nleqslv)
library(rootSolve)
```

# Problem 1


# Problem 2

# (a)

Let **$N(t)$** be the number of cars arriving at a parking lot by time **$t$**, according to the proposed scenario, we can model **$N(t)$** as a non-homogenous Poisson process. Such process has almost the same process as any other Poisson process, however, its rate is a function of time.

$N(t), t \in [0, \infty)$ is the non-homogenous Poisson process with rate $\lambda (t)$ where:

- $N(0) = 0$
- $N(t)$ has independent increments

We define 8:00 as $t=0$ with the following integrable function and each unit of $t$ equals to 1 hour:

$\lambda (t) = \begin{cases} 100 & 0 \leq t \leq \frac{1}{2} \\
600t - 200 & \frac{1}{2} \leq t \leq \frac{3}{4} \\
400t - 50 & \frac{3}{4} \leq t \leq 1 \\
-500t + 850 & 1 \leq t \leq 1.5 \\
\end{cases}$

So,

$E[N(t)] = \begin{cases} \end{cases}$

**lots of latex here**

# (b)

```{r, echo=FALSE, warning=FALSE, message=FALSE}

f <- function(l) {
    sum((exp(-l) * l^k)/(factorial(k))) - 0.8
}
k <- 0:150
nleqslv(0,f)
```
```{r, echo=FALSE, warning=FALSE, message=FALSE}

test <- 0
for (i in 1:149) {
    test <- test + (exp(-1.609438) * 1.609438^i)/(factorial(i))
}
test
```


```{r, echo=TRUE, warning=FALSE, message=FALSE}
f <- function(l) {
    1 - sum((exp(-l) * l^k)/(factorial(k))) - 0.8
}

tt = c()
aa = seq(0,100,0.1)
for (i in seq(0,100,0.1)) {
    tt = c(tt, f(i))
}
test <- data.frame(val=aa, tt=tt)

```

```{r, echo=TRUE, warning=FALSE, message=FALSE}
tt = c()
aa = seq(0,200,1)
for (i in aa) {
    tt = c(tt, ppois(149,lambda=i))
}
test <- data.frame(val=aa, tt=tt)
write.csv(test, 'test.csv')
```
```{r, echo=TRUE, warning=FALSE, message=FALSE}
tt = c()
aa = seq(0,140,1)
bb = seq(0, 30,1)
for (i in bb) {
    for (j in aa) {
        tt = c(tt, ppois(149,lambda=i))
    }
}
test <- data.frame(val=aa, tt=tt)
write.csv(test, 'test.csv')
```