---
title: 'Stochastic Processes: Assignment 1'
author: "Group 1: Javier Esteban Aragoneses, Mauricio Marcos Fajgenbaun, Danyu Zhang, Daniel Alonso"
date: "November 27th, 2020"
output: 'pdf_document'
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "./figures/"
)
```

Importing libraries

```{r, echo=FALSE, warning=FALSE, results='hide'}
library(markovchain)
library(matlib)
library(nleqslv)
library(rootSolve)
library(stringi)
```

# Problem 1

```{r, echo=FALSE, warning=FALSE, message=FALSE}
freq <- read.table('./data/Englishcharacters.txt') + 1
message <- paste(readLines("./data/messages.txt"))[1]

# decode function
decode <- function(message, freqs, iters) {
    # message separated by "" to iterate
    message <- strsplit(message, "")[[1]]

    # identity function
    fvals <- 1:27
    letters <- c("a", "b", "c", "d",
                "e", "f", "g", "h", 
                "i", "j", "k", "l", 
                "m", "n", "o", "p", 
                "q", "r", "s", "t", 
                "u", "v", "w", "x", 
                "y", "z", " ")
    names(fvals) <- letters
    f <- function(c) return(fvals[c])

    # score function
    score <- function(fun) {
        p <- 1
        for (i in 2:length(message)) {
            p <- p * freq[fun(message[i-1]),fun(message[i])]
        }
        return(p)
    }

    for (i in 1:iters) {
        # values for the function
        fvals_n <- 1:27

        # randomly choose 2 numbers and replace the 2 chosen vals in a copy of letters
        choice_1 <- as.integer(runif(1, min=1, max=27))
        choice_2 <- as.integer(runif(1, min=1, max=27))
        letters_n <- letters
        letters_n[choice_1] <- letters[choice_2]
        letters_n[choice_2] <- letters[choice_1]
        names(fvals_n) <- letters_n

        # f asterisk
        f_n <- function(c) return(fvals_n[c])

        scr <- (score(f_n)/score(f))
        print(scr)
        cond <- runif(1) <= scr
        rnd <- runif(1)
        if (is.na(cond) == TRUE) {
            for (l in 1:length(message)) {
                message[l] <- names(f(message[l]))
            }
        } else if (rnd >= scr){
            for (l in 1:length(message)) {
                message[l] <- names(f(message[l]))
            }
        } else if (rnd <= scr){
            f <- f_n
            letters <- letters_n
            names(fvals) <- letters
            for (l in 1:length(message)) {
                message[l] <- names(f(message[l]))
            }
        } 

    }
    return(stri_paste(message,collapse=''))
}
decode(message,freq,100)
```


# Problem 2

# (a)

Let **$N(t)$** be the number of cars arriving at a parking lot by time **$t$**, according to the proposed scenario, we can model **$N(t)$** as a non-homogenous Poisson process. Such process has almost the same process as any other Poisson process, however, its rate is a function of time.

$N(t), t \in [0, \infty)$ is the non-homogenous Poisson process with rate $\lambda (t)$ where:

- $N(0) = 0$
- $N(t)$ has independent increments

We define 8:00 as $t=0$ with the following integrable function and each unit of $t$ equals to 1 hour:

$\lambda (t) = \begin{cases} 100 & 0 \leq t \leq \frac{1}{2} \\
600t - 200 & \frac{1}{2} \leq t \leq \frac{3}{4} \\
400t - 50 & \frac{3}{4} \leq t \leq 1 \\
-500t + 850 & 1 \leq t \leq 1.5 \\
\end{cases}$

So,

$E[N(t)] = \begin{cases} \end{cases}$

**lots of latex here**

# (b)

```{r, echo=FALSE, warning=FALSE, message=FALSE}

f <- function(l) {
    sum((exp(-l) * l^k)/(factorial(k))) - 0.8
}
k <- 0:150
nleqslv(0,f)
```
```{r, echo=FALSE, warning=FALSE, message=FALSE}

test <- 0
for (i in 1:149) {
    test <- test + (exp(-1.609438) * 1.609438^i)/(factorial(i))
}
test
```


```{r, echo=TRUE, warning=FALSE, message=FALSE}
f <- function(l) {
    1 - sum((exp(-l) * l^k)/(factorial(k))) - 0.8
}

tt = c()
aa = seq(0,100,0.1)
for (i in seq(0,100,0.1)) {
    tt = c(tt, f(i))
}
test <- data.frame(val=aa, tt=tt)

```

```{r, echo=TRUE, warning=FALSE, message=FALSE}
tt = c()
aa = seq(0,200,1)
for (i in aa) {
    tt = c(tt, ppois(149,lambda=i))
}
test <- data.frame(val=aa, tt=tt)
write.csv(test, 'test.csv')
```
```{r, echo=TRUE, warning=FALSE, message=FALSE}
tt = c()
aa = seq(0,140,1)
bb = seq(0, 30,1)
for (i in bb) {
    for (j in aa) {
        tt = c(tt, ppois(149,lambda=i))
    }
}
test <- data.frame(val=aa, tt=tt)
write.csv(test, 'test.csv')
```